\section*{Contribute}

This manuscript is far from perfect: 
%
Interesting aspects are omitted, because I %the authors
didn't have the time, knowledge, or expertise to add them. For instance, the specification 
hasn't been entirely covered, as is documented 
(\href{https://github.com/BenSmyth/tls-tutorial/blob/master/rfc8446-annotated.txt}{rfc8446-annotated.txt}); 
discussion of security guarantees are notably lacking; and an introduction to the underlying 
cryptography is absent. (E.g., some details on DHKE, AEAD, etc.\ would be grand.) 
%
Directions for further exploration are missing, hands-on teaching opportunities foregone. For 
instance, a Davies-style exploration of TLS on-the-wire, with notes on Wireshark and 
\code{SSLKEYLOGFILE} -- perhaps as dirty as readers can get, without bursting out soldering 
irons. (Cf. Davies's \emph{Implementing SSL/TLS: Using Cryptography and PKI}.)
%
Mistakes and issues are no doubt numerous. (Some are flagged by tokens \texttt{\textbackslash{}ifSpecNotes}, 
\texttt{\textbackslash{}ifImplNotes}, and \texttt{\textbackslash{}ifPresentationNotes}.)


%Publication is antiquated; manuscripts can continually evolve: 
\paragraph{Publication is antiquated; help evolve this manuscript:}
I %The authors
encourage \emph{you} to improve this manuscript. Fix a typo. Patch grammar.
Revise awkward, overcomplicated, or otherwise poorly-written passages. 
Contribute an entire section. Help evolve this manuscript:
%
\begin{center}
  \url{https://github.com/BenSmyth/tls-tutorial/}
\end{center}
%
(Perhaps get in touch prior to writing an entire section! We should probably reach consensus 
on what to add.) Contributions will be recognised through acknowledgements or co-authorship.


